Indexed Containers

### 2011-05-09 journal/paper.tex
first version of introduction

\usepackage{url}
\usepackage{times}
\usepackage{amsmath}
\usepackage{xypic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{stmaryrd}

\newcommand{\note}[1]{}

@@

Blah

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{ℕ}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\Conid{ℕ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;(\Varid{n}\;\in\;\Conid{ℕ})\;\rightarrow\;\Conid{ℕ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{List}\;(\Conid{A}\;\in\;\Conid{Set})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}[\mskip1.5mu \mskip1.5mu]\;{}\<[8]%
\>[8]{}\in\;{}\<[24]%
\>[24]{}\Conid{List}\;\Conid{A}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{\char95 ∷\char95 }\;{}\<[8]%
\>[8]{}\in\;\Conid{A}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;{}\<[24]%
\>[24]{}\Conid{List}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Lam}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[8]%
\>[8]{}\in\;(\Varid{n}\;\in\;\Conid{ℕ})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{app}\;{}\<[8]%
\>[8]{}\in\;(\Varid{f}\;\Varid{a}\;\in\;\Conid{Lam})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lam}\;{}\<[8]%
\>[8]{}\in\;(\Varid{t}\;\in\;\Conid{Lam})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Fin}\;\in\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Varid{n}\mskip1.5mu\}\;{}\<[30]%
\>[30]{}\rightarrow\;\Conid{Fin}\;(\Varid{suc}\;\Varid{n}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Varid{n}\mskip1.5mu\}\;(\Varid{i}\;\in\;\Conid{Fin}\;\Varid{n})\;{}\<[30]%
\>[30]{}\rightarrow\;\Conid{Fin}\;(\Varid{suc}\;\Varid{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

@@

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{mutual}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{data}\;\Conid{NeLam}\;(\Varid{n}\;\in\;\Conid{ℕ})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{var}\;{}\<[10]%
\>[10]{}\in\;(\Varid{i}\;\in\;\Conid{Fin}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NeLam}\;\Varid{n}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{app}\;{}\<[10]%
\>[10]{}\in\;(\Varid{f}\;\in\;\Conid{NeLam}\;\Varid{n})\;(\Varid{a}\;\in\;\Conid{NfLam}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NeLam}\;\Varid{n}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{data}\;\Conid{NfLam}\;(\Varid{n}\;\in\;\Conid{ℕ})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{lam}\;{}\<[10]%
\>[10]{}\in\;(\Varid{t}\;\in\;\Conid{NfLam}\;(\Varid{suc}\;\Varid{n}))\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NfLam}\;\Varid{n}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ne}\;{}\<[10]%
\>[10]{}\in\;(\Varid{t}\;\in\;\Conid{NeLam}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NfLam}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Ty}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\iota\;{}\<[6]%
\>[6]{}\in\;\Conid{Ty}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\anonymous \Rightarrow\anonymous \;{}\<[11]%
\>[11]{}\in\;(\sigma\;\tau\;\in\;\Conid{Ty})\;\rightarrow\;\Conid{Ty}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{Var}\;(\tau\;\in\;\Conid{Ty})\;\in\;\Conid{List}\;\Conid{Ty}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Gamma\mskip1.5mu\}\;{}\<[34]%
\>[34]{}\rightarrow\;\Conid{Var}\;\tau\;(\tau\;\Varid{∷}\;\Gamma){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\Gamma\mskip1.5mu\}\;(\Varid{i}\;\in\;\Conid{Var}\;\tau\;\Gamma)\;{}\<[34]%
\>[34]{}\rightarrow\;\Conid{Var}\;\tau\;(\sigma\;\Varid{∷}\;\Gamma){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{STLam}\;(\Gamma\;\in\;\Conid{List}\;\Conid{Ty})\;\in\;\Conid{Ty}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{i}\;\in\;\Conid{Var}\;\tau\;\Gamma)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;\tau{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{app}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{f}\;\in\;\Conid{STLam}\;\Gamma\;(\sigma\;\Rightarrow\;\tau))\;{}\<[E]%
\\
\>[19]{}(\Varid{a}\;\in\;\Conid{STLam}\;\Gamma\;\sigma)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;\tau{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lam}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{t}\;\in\;\Conid{STLam}\;(\sigma\;\Varid{∷}\;\Gamma)\;\tau)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;(\sigma\;\Rightarrow\;\tau){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\section{Background}

>>>>

\usepackage{url}
\usepackage{times}
\usepackage{amsmath}
\usepackage{xypic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{color}

\newcommand{\note}[1]{}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{Todo:~}#1}}

@@

  We show that the syntactically rich notion of inductive families can
  be reduced to a core type theory with a fixed number of type
  constructors exploiting the novel notion of indexed containers.
  Indexed containers generalize simple containers, capturing strictly
  positive families instead of just strictly positive types, without
  having to extend the core type theory. Other applications of indexed
  containers include datatype-generic programming and reasoning about
  polymorphic functions. The construction presented here has been
  formalized using the Agda system.

@@

\noindent Inductive datatypes are a central feature of modern Type Theory
(e.g. COQ~\cite{CIC}) or functional programming (e.g. 
Haskell\footnote{Here we shall view Haskell as an approximation of strong
  functional programming as proposed by Turner \cite{sfp} and ignore
non-termination.})

Examples include the natural numbers al Peano:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{ℕ}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\Conid{ℕ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;(\Varid{n}\;\in\;\Conid{ℕ})\;\rightarrow\;\Conid{ℕ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
the set of lists indexed by any given set:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{List}\;(\Conid{A}\;\in\;\Conid{Set})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}[\mskip1.5mu \mskip1.5mu]\;{}\<[8]%
\>[8]{}\in\;{}\<[24]%
\>[24]{}\Conid{List}\;\Conid{A}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{\char95 ∷\char95 }\;{}\<[8]%
\>[8]{}\in\;\Conid{A}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;{}\<[24]%
\>[24]{}\Conid{List}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
and the set of  de Bruijn $\lambda$-terms:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Lam}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[8]%
\>[8]{}\in\;(\Varid{n}\;\in\;\Conid{ℕ})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{app}\;{}\<[8]%
\>[8]{}\in\;(\Varid{f}\;\Varid{a}\;\in\;\Conid{Lam})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lam}\;{}\<[8]%
\>[8]{}\in\;(\Varid{t}\;\in\;\Conid{Lam})\;{}\<[23]%
\>[23]{}\rightarrow\;\Conid{Lam}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\noindent An elegant way to formalize and reason about inductive types
is to model them as the initial algebra of an endofunctor, we can
define the siganture functors cirresponding to each of the examples above:

@@

This perspective has been very
successful in providing a generic approach to programming with and
reasoning about inductive types, e.g. see the \emph{Algebra of
Programming} \cite{BirdDeMoor:AlgProp}.

While the theory of inductive types is well developed, we often want
to have a finer, more expressive, notion of types, for example to
ensure the absence of run time errors --- access to arrays out of
range or access to an undefined variable in the previous example of
$\lambda$-terms.

To model this we move to the notion of an inductive
family in Type Theory. A family is a type indexed by another already
given type. The first example is the family of finite sets \ensuremath{\Conid{Fin}} which
assigns to any natural number \ensuremath{\Varid{n}} a set \ensuremath{\Conid{Fin}\;\Varid{n}} which has exactly
\ensuremath{\Varid{n}}-elements. \ensuremath{\Conid{Fin}} can be used where in conventional reasoning we
assume any finite set, e.g. when dealing with a finite address apce or
a finite set of variables. The inductive definition of \ensuremath{\Conid{Fin}} refines
the type of natural numbers:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Fin}\;\in\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Varid{n}\mskip1.5mu\}\;{}\<[30]%
\>[30]{}\rightarrow\;\Conid{Fin}\;(\Varid{suc}\;\Varid{n}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Varid{n}\mskip1.5mu\}\;(\Varid{i}\;\in\;\Conid{Fin}\;\Varid{n})\;{}\<[30]%
\>[30]{}\rightarrow\;\Conid{Fin}\;(\Varid{suc}\;\Varid{n}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the same fashion we can refine the type of lists to the type of
vectors which are additionally indexed by a number indicating the
length of the vector:

@@

Using \ensuremath{\Conid{Fin}} and \ensuremath{\Conid{Vec}} instead of \ensuremath{\Conid{Nat}} and \ensuremath{\Conid{List}} enables us to write
a total projection function projecting the nth element out of vector:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Maybe}\;\Conid{A}{}\<[E]%
\\
\>[B]{}[\mskip1.5mu \mskip1.5mu]\;\mathbin{!!}\;\Varid{n}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{nothing}{}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{zero}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{just}\;\Varid{a}{}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{suc}\;\Varid{n}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{as}\;\mathbin{!!}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Note, that a corresponding function \ensuremath{\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{A}} is not definable in a total langauge like Agda.

Finally we can define the notion of a well-scoped lambda term with
\ensuremath{\Conid{ScLam}} which assigns to a natural number \ensuremath{\Varid{n}} the set of $\lambda$-terms
with at most \ensuremath{\Varid{n}} free variables \ensuremath{\Conid{ScLam}\;\Varid{n}}. DeBruijn variables are now
modelled by elements of \ensuremath{\Conid{Fin}\;\Varid{n}} replacing \ensuremath{\Conid{Nat}} in the previous,
unindexed definition of $\lambda$-terms \ensuremath{\Conid{Lam}}.

@@

\noindent
Importantly, the constructor
\ensuremath{\Varid{lam}} reduces the number of \emph{free} variables by one --- by
binding one. 
Inductive families may be mutually defined, for example the scoped
versions of  $\beta$ (\ensuremath{\Conid{NfLam}})
normal forms and neutral $\lambda$-terms (\ensuremath{\Conid{NeLam}}): 
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{mutual}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{data}\;\Conid{NeLam}\;(\Varid{n}\;\in\;\Conid{ℕ})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{var}\;{}\<[10]%
\>[10]{}\in\;(\Varid{i}\;\in\;\Conid{Fin}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NeLam}\;\Varid{n}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{app}\;{}\<[10]%
\>[10]{}\in\;(\Varid{f}\;\in\;\Conid{NeLam}\;\Varid{n})\;(\Varid{a}\;\in\;\Conid{NfLam}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NeLam}\;\Varid{n}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{data}\;\Conid{NfLam}\;(\Varid{n}\;\in\;\Conid{ℕ})\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{lam}\;{}\<[10]%
\>[10]{}\in\;(\Varid{t}\;\in\;\Conid{NfLam}\;(\Varid{suc}\;\Varid{n}))\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NfLam}\;\Varid{n}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ne}\;{}\<[10]%
\>[10]{}\in\;(\Varid{t}\;\in\;\Conid{NeLam}\;\Varid{n})\;{}\<[41]%
\>[41]{}\rightarrow\;\Conid{NfLam}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The initial algebra semantics of inductive types can be extended to
model inductive families by replacing functors on the category \ensuremath{\Conid{Set}}
with functors on the category of families indexed by a given type - in
the case of $\lambda$-terms this indexing type was \ensuremath{\Conid{Nat}}. The objects
of the category of families indexed over a type \ensuremath{\Conid{I}\;\in\;\Varid{set}} are
\ensuremath{\Conid{I}}-indexed families, i.e. functions of type \ensuremath{\Conid{I}\;\rightarrow\;\Conid{Set}}, and a
morphism between \ensuremath{\Conid{I}}-indexed families \ensuremath{\Conid{A},\Conid{B}\;\in\;\Conid{I}\;\rightarrow\;\Conid{Set}} is given by a
family of maps \ensuremath{\Varid{f}\;\in\;(\Varid{i}\;\in\;\Conid{I})\;\to \;\Conid{A}\;\Varid{i}\;\to \;\Conid{B}\;\Varid{i}}
Indeed, this category
is easily seen to be isomorphic to the slice category $\ensuremath{\Conid{Set}}/ \ensuremath{\Conid{I}}$ but
the chosen representation is more convenient type-theoretically. Using
$\Sigma$-types and equality types from Type Theory, we can define the
following endofunctors \ensuremath{\Conid{F}_{\Conid{Fin}},\Conid{F}_{\Conid{Vec}}} and \ensuremath{\Conid{F}_{\Conid{Lam}}}
on the category of families
over \ensuremath{\Conid{Nat}} whose initial algebras are \ensuremath{\Conid{Fin}} and \ensuremath{\Conid{Lam}}, respectively:

@@

The equality type expresses the focussed character of the
constructors for \ensuremath{\Conid{Fin}}.

\todo{Discuss mutual case.}

This approach extends uniformly to more complicated examples such as
the family of typed $\lambda$-terms, using lists of types  to
represent typing contexts:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Ty}\;\in\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\iota\;{}\<[6]%
\>[6]{}\in\;\Conid{Ty}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\anonymous \Rightarrow\anonymous \;{}\<[11]%
\>[11]{}\in\;(\sigma\;\tau\;\in\;\Conid{Ty})\;\rightarrow\;\Conid{Ty}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{Var}\;(\tau\;\in\;\Conid{Ty})\;\in\;\Conid{List}\;\Conid{Ty}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{zero}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \Gamma\mskip1.5mu\}\;{}\<[34]%
\>[34]{}\rightarrow\;\Conid{Var}\;\tau\;(\tau\;\Varid{∷}\;\Gamma){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{suc}\;{}\<[9]%
\>[9]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\Gamma\mskip1.5mu\}\;(\Varid{i}\;\in\;\Conid{Var}\;\tau\;\Gamma)\;{}\<[34]%
\>[34]{}\rightarrow\;\Conid{Var}\;\tau\;(\sigma\;\Varid{∷}\;\Gamma){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{STLam}\;(\Gamma\;\in\;\Conid{List}\;\Conid{Ty})\;\in\;\Conid{Ty}\;\rightarrow\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{i}\;\in\;\Conid{Var}\;\tau\;\Gamma)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;\tau{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{app}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{f}\;\in\;\Conid{STLam}\;\Gamma\;(\sigma\;\Rightarrow\;\tau))\;{}\<[E]%
\\
\>[19]{}(\Varid{a}\;\in\;\Conid{STLam}\;\Gamma\;\sigma)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;\tau{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lam}\;{}\<[8]%
\>[8]{}\in\;\forall{}\;\{\mskip1.5mu \sigma\;\tau\mskip1.5mu\}\;{}\<[19]%
\>[19]{}(\Varid{t}\;\in\;\Conid{STLam}\;(\sigma\;\Varid{∷}\;\Gamma)\;\tau)\;{}\<[44]%
\>[44]{}\rightarrow\;\Conid{STLam}\;\Gamma\;(\sigma\;\Rightarrow\;\tau){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\noindent
Types like this can be used to implement a tag-free, terminating
evaluator~\cite{bsn}. To obtain the corresponding functors
is a laborious but straightforward exercise.

\todo{Expand here?}

\subsection{Overview over the paper}
\label{sec:overview-over-paper}

We develop our type theoretic and categorical background in section
\ref{sec:background} and also summarize the basic definitions of
non-indexed containers. In section \ref{sec:ifunc} we develop the
concept of an indexed functor, showing that this is a relative monad
and presenting basic constructions on indexed functors including the
definition of a parametrized initial algebra. In section
\ref{sec:icont} we devlop the basic theory of indexed containers and
relate them to indexed functors. Subsequently in section
\ref{sec:initalg} we construct initial algebras of indexed containers
assuming the existence of indexed W-types, this can be dualized to
showing the existence of terminal coalgebras from indexed M-types 
\ref{sec:termcoalg}. Both requirements, indexed W-types and indexed
M-types can be derived from ordinary W-types, this is shown in section
\ref{sec:w-enough}. Finally, we define a syntax from strictly positive
families and interpret this using indexed containers \ref{sec:spf}.

\subsection{Related work}
\label{sec:related-work}

\noindent
Inductive families are the backbone of
dependently typed programming as present in Epigram or
Agda~\cite{Agda}. Coq also supports the definition of inductive families
but programming with them is rather hard --- a situation which has been
improved by the new \texttt{Program} tactic~\cite{sozeau}. 
More recently, the implementation of Generalized Algebraic Datatypes 
(GADTs)~\cite{Hinze:GADT} 
allows \ensuremath{\Conid{Fin}} and \ensuremath{\Conid{Lam}} to be encoded in Haskell:
\begin{tabbing}\tt
~data~Fin~a~where~\\
\tt ~~~FZero~\char58{}\char58{}~Fin~\char40{}Maybe~a\char41{}\\
\tt ~~~FSucc~\char58{}\char58{}~Fin~a~\char45{}\char62{}~Fin~\char40{}Maybe~a\char41{}\\
\tt ~\\
\tt ~data~Lam~a~where~\\
\tt ~~~Var~\char58{}\char58{}~Fin~a~\char45{}\char62{}~Lam~a\\
\tt ~~~App~\char58{}\char58{}~Lam~a~\char45{}\char62{}~Lam~a~\char45{}\char62{}~Lam~a\\
\tt ~~~Abs~\char58{}\char58{}~Lam~\char40{}Maybe~a\char41{}~\char45{}\char62{}~Lam~a
\end{tabbing}
Here \texttt{Fin} and \texttt{Lam} are indexed by types instead of
natural numbers; The type constructor \texttt{Maybe} serves as a type level
copy of the $\succ$ constructor for natural numbers.
Note that \texttt{Lam} is actually just a nested datatype 
\cite{alti:csl99} while \texttt{Fin} exploits the full power of
GADTs because the range of the constructors is constrained.

\todo{Rewrite and expand}
